Impact of Federal Party Leadership Changes on Donation Patterns

### 2024-12-03 paper/paper.qmd
updated data section to add tables and graphs

The paper is organized as follows: Section \ref{sec-data} discusses the data sources and methodology, followed by an explanation of the variables used in the analysis. The regression models and results are then presented, along with interpretations. The paper concludes by discussing the broader implications of the findings and suggesting directions for future research.[NEED TO UPDATE THE REFERENCES]

# Data {#sec-data}

@@

We use the statistical programming language R [@citeR].... Our data [@shelter].... Following @tellingstories, we consider...

Overview text

@@

Some paragraphs about how we go from a phenomena in the world to an entry in the dataset.

## Outcome variables

Add graphs, tables and text. Use sub-sub-headings for each outcome variable or update the subheading to be singular.

Some of our data is of penguins (@fig-bills), from @palmerpenguins.

```{r}
#| label: fig-bills
#| fig-cap: Bills of penguins
#| echo: false

ggplot(penguins, aes(x = island, fill = species)) +
  geom_bar(alpha = 0.8) +
  scale_fill_manual(values = c("darkorange","purple","cyan4"),
                    guide = "none") +
  theme_minimal() +
  facet_wrap(~species, ncol = 1) +
  coord_flip()
```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}
#| label: fig-planes
#| fig-cap: Relationship between wing length and width
#| echo: false
#| warning: false
#| message: false

analysis_data <- read_csv(here::here("data/02-analysis_data/analysis_data.csv"))

analysis_data |> 
  ggplot(aes(x = width, y = length)) +
  geom_point(alpha = 0.8) +
  theme_minimal() +
  labs(x = "Wing width (mm)",
       y = "Wing length (mm)")
```

Talk way more about it.

## Predictor variables

Add graphs, tables and text.

Use sub-sub-headings for each outcome variable and feel free to combine a few into one if they go together naturally.

@@

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

@@

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

@@

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

>>>>

The paper is organized as follows: Section \ref{sec-data} discusses the
data sources and methodology, followed by an explanation of the
variables used in the analysis. The regression models and results are
then presented, along with interpretations. The paper concludes by
discussing the broader implications of the findings and suggesting
directions for future research.\[NEED TO UPDATE THE REFERENCES\]

# Data

```{r echo=FALSE, message=FALSE, warning=FALSE}
#### Workspace setup ####
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(readr)   
library(tidyr)

#### Read data ####
analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")

# Separate data into Ontario and Federal datasets
ontario_data <- analysis_data %>% filter(region == "Ontario")
federal_data <- analysis_data %>% filter(region == "Federal")
```

@@

The dataset used in this analysis comprises political donation records spanning from 2013 to 2024, covering both the federal level and the province of Ontario in Canada. These records were sourced from the Investigative Journalism Foundation (IJF), which systematically collects and maintains comprehensive data on political donations across various Canadian jurisdictions. The IJF gathers daily updates by monitoring elections agency websites at the federal level and within each province and territory, ensuring the dataset remains current and robust. Although the IJF's repository includes historical data dating back to 1993, this study focuses exclusively on the 2013–2024 period to capture recent trends and changes in political financing dynamics (Investigative Journalism Foundation 2024).

The analysis was conducted using the statistical programming language R (R Core Team 2024) and several packages for data manipulation and visualization, including `dplyr` (Wickham et al. 2023), `ggplot2` (Wickham 2016), `readr` (Wickham et al. 2023), and `tidyr` (Wickham et al. 2023).

Political donations in Canada are subject to legal disclosure requirements, promoting transparency and accountability within the democratic process. Both federal and provincial election finance laws mandate that political parties, candidates, and associated entities report donations exceeding specific thresholds. These disclosures vary by jurisdiction in terms of frequency and reporting formats. For instance, Ontario requires annual returns for general and by-elections, leadership races, and donations to parties, candidates, and constituency associations, all available in downloadable spreadsheet formats from 2007 onward. This regulatory framework ensures that significant contributions are systematically documented and publicly accessible, providing a reliable foundation for this analysis.

@@

The primary objective of this analysis is to investigate the impact of a political party's power status on donation patterns within Canadian federal and Ontario provincial elections. The dataset includes detailed records of donations made to various recipients, such as political parties, party leadership contestants, riding associations (also known as electoral district or constituency associations), and individual riding candidates. Donations are tracked across different types of electoral events, including general elections, by-elections, and leadership races.

Key variables within the dataset are as follows:

- **Political Party:** A categorical variable indicating the recipient political party, encompassing both major parties (e.g., Liberal Party, Conservative Party, New Democratic Party) and minor parties.

- **Donation Year:** A numerical variable representing the year the donation was made, ranging from 2013 to 2024. This captures different political climates and election periods.

- **Amount Donated:** A continuous variable denoting the monetary value of each donation, varying widely due to differing donor capacities and legal contribution limits.

### Constructed Variables

Several variables were constructed to facilitate a more nuanced analysis:

- **Recipient_in_Power:** A binary variable indicating whether the recipient party was in power during the donation year (1 for in power, 0 for not in power). This was determined by cross-referencing the donation year with historical records of party leadership at both the federal and provincial levels.

- **Party_Size:** A binary variable categorizing parties as major (1) or minor (0), based on their representation in the legislature. Major parties are those with significant representation or historical influence.

- **Election_Year:** A binary variable indicating whether the donation was made during an election year (1) or not (0), based on official election dates at the federal and provincial levels.

## Data Cleaning

The raw donation records underwent several preprocessing steps to ensure the dataset's integrity and suitability for analysis:

1. **Handling Missing Values:** Records with missing or incomplete information, particularly in key variables such as donation amount or recipient party, were excluded from the analysis to maintain data quality.

2. **Standardizing Party Names:** Political party names were standardized to ensure consistency. For example, variations like "Progressive Conservative Party of Ontario" and "PC Party of Ontario" were consolidated under a single standardized name.

3. **Removing Outliers:** Extreme donation amounts that could skew the analysis were identified and removed based on predefined thresholds, acknowledging the legal donation limits in Canada.

4. **Aggregating Donations:** Donations were aggregated to reflect total amounts per party per year, facilitating a higher-level analysis of donation patterns.

5. **Creating Binary Variables:** Binary indicators were generated for key variables such as `Recipient_in_Power`, `Party_Size`, and `Election_Year` to streamline regression analysis.

Additional details on the data cleaning process, including code snippets and specific handling of anomalies, are provided in Appendix A.

## Outcome Variables

The main outcome variable of interest is the **Amount Donated**, representing the monetary value of each donation made to a political party within a given year. This continuous variable is critical for assessing the financial support received by parties and understanding how it correlates with factors like power status and party size.

The distribution of donation amounts varies between the federal and provincial levels. At the federal level, the average donation amount is higher, reflecting different legal contribution limits and donor behaviors.

## Predictor Variables

Several key predictor variables were identified to evaluate factors influencing donation amounts:

- **Recipient_in_Power:** Indicates whether the recipient party was in power during the donation year (1) or not (0). This variable is central to assessing the impact of political power on donation patterns.

- **Party_Size:** Categorizes parties as major (1) or minor (0). Major parties are expected to receive more donations due to greater visibility and established support bases.

- **Election_Year:** Indicates whether the donation was made during an election year (1) or not (0). Election years often see increased fundraising activities.

### Additional Predictor Variables

- **Log_Total_Donations:** The logarithm of total donations received by a party in a given year, calculated as `log(Total_Donations + 1)` to handle zero values and normalize the distribution.

- **Interaction Terms:** Interaction terms such as `Recipient_in_Power * Party_Size` were included in the models to explore how the effect of being in power varies between major and minor parties.

## Summary Statistics

### Donations Based on Recipient Power Status

#### Ontario

Table 1 presents the total donations and average donation amounts in Ontario, categorized by whether the recipient party was in power.

**Table 1: Donations by Power Status in Ontario**
```{r echo=FALSE, message=FALSE, warning=FALSE}
## Donations Based on Recipient Power Status ##
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

Figure 1 illustrates the total donations and average donation amounts over the years in Ontario.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Total and Average Donations Over Years - Ontario
donations_by_year_ontario <- ontario_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

library(ggplot2)
library(gridExtra)

p1_ontario <- ggplot(donations_by_year_ontario, aes(x = donation_year, y = total_amount)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(title = "Total Donations Over Years - Ontario",
       x = "Year", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

p2_ontario <- ggplot(donations_by_year_ontario, aes(x = donation_year, y = avg_amount)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(title = "Average Donation Amount Over Years - Ontario",
       x = "Year", y = "Average Donation Amount ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

grid.arrange(p1_ontario, p2_ontario, ncol = 2)
```

*Figure 1: Total and Average Donations Over Years in Ontario*

Figure 3 displays the total donations by political party in Ontario.

```{r echo=FALSE, message=FALSE, warning=FALSE}
donations_by_party_ontario <- ontario_data %>%
  group_by(political_party) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

#Plotting Donations by Political Party - Ontario

p3_ontario <- ggplot(donations_by_party_ontario, aes(x = reorder(political_party, -total_amount), y = total_amount, fill = political_party)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Donations by Political Party - Ontario",
       x = "Political Party", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(p3_ontario)
```

*Figure 3: Total Donations by Political Party in Ontario*

#### Federal

Similarly, Table 2 shows the donations at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
## Donations Based on Recipient Power Status ##
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

Similarly, Figure 2 shows the total and average donations over the years at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Total and Average Donations Over Years - Federal
donations_by_year_federal <- federal_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

p1_federal <- ggplot(donations_by_year_federal, aes(x = donation_year, y = total_amount)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue") +
  labs(title = "Total Donations Over Years - Federal",
       x = "Year", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

p2_federal <- ggplot(donations_by_year_federal, aes(x = donation_year, y = avg_amount)) +
  geom_line(color = "darkgreen") +
  geom_point(color = "darkgreen") +
  labs(title = "Average Donation Amount Over Years - Federal",
       x = "Year", y = "Average Donation Amount ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal()

grid.arrange(p1_federal, p2_federal, ncol = 2)
```

*Figure 2: Total and Average Donations Over Years at Federal Level*

Figure 4 shows the total donations by political party at the federal level.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting Donations by Political Party - Federal
donations_by_party_federal <- federal_data %>%
  group_by(political_party) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

p3_federal <- ggplot(donations_by_party_federal, aes(x = reorder(political_party, -total_amount), y = total_amount, fill = political_party)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Donations by Political Party - Federal",
       x = "Political Party", y = "Total Amount Donated ($)") +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

print(p3_federal)
```

*Figure 4: Total Donations by Political Party at Federal Level*

## Relationships Between Variables

Preliminary analysis indicates that political parties not in power receive higher total donations on average compared to those in power, at both federal and provincial levels. This trend suggests that donors may be motivated to support opposition parties to influence political change.

Additionally, major parties consistently attract more donations than minor parties, highlighting the importance of party prominence in fundraising efforts.

@@

@@

@@
